\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[]{amsmath}
\usepackage{amsthm,amssymb}
\usepackage{enumerate}% http://ctan.org/pkg/enumerate
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{ mathrsfs}
\usepackage{mathtools}
\usepackage[margin=0.5in]{geometry}

\usepackage{tikz}
\usepackage{pgfplots}

% no indents pls
\setlength\parindent{0pt}

\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}

%%% defs for theorem environments
\renewcommand{\qedsymbol}{\rule{0.7em}{0.7em}} % black box for QED
\newtheorem*{thm}{Theorem}
\newtheorem{lem}{Lemma}
% defining style for Cases
\newtheoremstyle{casestyle}{\topsep}{1pt}{}{0pt}{\itshape}{.}{5pt plus 1pt minus 1pt}{}
\theoremstyle{casestyle}
\newtheorem{case}{Case}
% smallcaps-ing "Proof" header
\let\oldproofname=\proofname
\renewcommand{\proofname}{\rm\sc{\oldproofname}}

\begin{document}

\section *{Problem 2}

We can construct a binomial heap given an order $k$ and minimum number of nodes $m$ by inserting $2^{\ceil{\log_2 (m+1)}}$ nodes and then calling extract-min once to consolidate the trees. This results in a heap with $2^{\ceil{\log_2 (m+1)}}-1$ nodes that has child subtrees of order $0 \dots \ceil{\log_2 (m+1)}-1$. Since the number of nodes is $2^{\ceil{\log_2 (m+1)}-1} \geq m$ nodes. We can then find the minimum key of the tree, $s$ and call decrease-key on the roots of the child trees starting from the tree of order $0$ for $k-1$ times, setting the root of the tree to $s-1$. Since $s-1$ is guaranteed to be lower than the root of the tree, this will cause the children binomial heaps to be cut off and increase the order of the overall binomial heap by one every time. We can repeat this $k-1$ times until we have a binomial heap of order $k$ with more than $m$ nodes, as desired.

% We can show that there is no upper bound on the number of nodes in the tree with a fixed $k$ via induction. Given a Fibonacci heap of any size with an order $k$ and number of nodes $n$, we can create another Fibonacci heap of order $k$ containing at least $n+1$ nodes. \\\\

% Base Case. Given any $k$, we can construct a Fibonacci heap of order $k$ by inserting $2^k$ nodes. After constructing the heap of desired order, we call extract-min once to consolidate branches, giving $2^k-1$ nodes and one child tree of each order from $0$ to $k-1$. \\\\

% Inductive Step. We will show that given a tree of order $k$ with $n$ nodes and child binomial trees of order $i$ through $i+k-1$, we can perform operations on it to create a tree with at least $n+1$ nodes of order $j$ through $j+k-1$. \\ We can achieve this by taking the 



% We can find the value of the smallest node in the tree $s$ and decrease the key of any non-root node to $s-1$ (if there is only one node in the tree, we can move directly to phase 2, treating the single node as if it were ``cut off''). This will cause at least one node to be cut off of the tree since there is no node with a value lower than $s$ before the decrease operation (i.e. its parent must have a value greater than $s-1$). \\\\

% Phase 2.

% We then insert a 

\newpage


\section *{Problem 3}

We begin with a lazy binomial heap and perform modifications in order to achieve our desired  amortized runtime of new-pq, insert, find-min, meld, and add-to-all in amortized time $O(1)$ and extract-min in $O(\log n)$. We will design around the function add-to-all since it is the major addition to our data structure.\\

In order to achieve amortized constant time melds, we also introduce a separate, auxiliary binomial heap that will keep track of the melded queues and the minimum elements of those queues at all times. For clarity, we will refer to this auxiliary heap as $A$ and the overall structure that we are building as $B$. The nodes in $A$ will have keys representing the value of the minimum element of each queue that was melded to $B$ and values that are pointers to the elements themselves. Operations on $A$ are standard lazy binomial heap operations as outlined in lecture.\\

We introduce the notion of an augmented property $d$ on each node of $B$ that starts initially at $0$. This field is an integer that represents the add-to-all $\Delta k$ values that have been added to the priority of the node and all its children over its lifetime. In addition, we keep a counter $c$ with an initial value of 0 in $B$ to keep track of add-to-all's that have been performed to $B$ overall. For example, a node (with no parents) with $d=3$ and priority $2$ has an effective priority of $3+2+2=7$ if $c$ is $2$ for the queue that it is in and its children's priorities are also effectively $d+c=3+2$ higher. \\

These operations assume that they are being performed on a data structure $B$ with auxiliary heap $A$.\\\\
\textbf{new-pq}: Initialize an empty binomial heap for $A$ and $B$. No nodes are on either heap. \\\\
\textbf{insert}$(v,k)$: Create a node $n_1$ with value $v$, key $k$, and aux value $d=-c$ (since the effective priority of every element in $B$ is actually $k+d+c$). Append this element to the end of the linked list of roots in $B$. Update the min pointer (as in a normal lazy binomial heap) for $B$ if the value $k+d$ of the new node is less than the $k+d$ node that the min pointer is pointing at.\\\\
\textbf{find-min}: Perform find-min on $A$ to get the node with the lowest key, $n$. Compare this key to the $k+d$ of the node pointed to by the min pointer and return the node with the lower value..\\\\
\textbf{meld}$(B,pq_2)$: Insert the minimum element of $pq_2$ (using find-min to get) into $A$. Subtract $c$ from $pq_2.c$ to account for the counter. Return $B$ as the new tree.\\\\
\textbf{add-to-all}$(s)$: Set $c$ to $c + s$.  \\\\
\textbf{extract-min}: \\ Phase 1: Extraction. Find the minimum element $n_a$ in $A$ and compare it to the node pointed to by the min pointer (as in find-min). Let this node be $n$. Take the root trees of the queue that contains $n$ (this information is stored in $A$ as well) and add $n.d$ to the children's auxiliary values ($d$-values) (i.e. increment/decrement the axuiliary priority of the children by the auxiliary priority of the root). Take these children and append them to the linked list of root nodes in $B$. For each of the melded queues in $A$, add its root trees to the linked list in $B$ as well (excluding the tree that has $n$ as the root if $n$ was taken from $A$). $A$ can then be cleared. \\\\ Phase 2: Merge. We can now merge the trees that are in the linked list of $B$ as demonstated in lecture 8: by creating space for $O(\log n)$ trees and then assigning and merging trees based on their size (removing them from the linked list), a standard lazy binomial operation. When the merging in the binomial heap occurs, we have two trees: $T_1$ and $T_2$ that must be merged. Without loss of generality, let $T_1$ have a lower $k+d$ value. We must maintain accuracy of the $d$ fields stored in the node, so we set $T_2.d = T_2.d - T_1.d$ and let $T_1$ be the root of the new merged tree. This occurs on every tree merge operation to preserve the accuracy of $d$ at the root. At the end of \textbf{extract-min} and its merging, there will be only $O(\log n)$ trees.\\

This series of operations allows us to maintain the node with the lowest priority at the top of the heap while offering operations such as add-to-all and meld in amortized constant time. Since the counter values and auxiliary $d$ values are propagated to the children every time the minimum node is extracted/replaced, we maintain correct functionality. Since everything in $B$ is subject to a shift by $c$, we can do add-to-all easily as well. \\\\

\begin{proof} of Runtime
In order to prove amortized runtime, we use a potential function $\phi=\text{\# nodes in } A + \text{\# trees in } B$. Let $T_A$ be the number of nodes in $A$ and $T_B$ be the number of root trees in $B$. The potential function thus becomes $\phi=T_A+T_B$  \\
\textbf{new-pq}: $O(1)$ actual. No potential change. $O(1)$.\\\\
\textbf{insert}$(v,k)$: $O(1)$ actual (adding to linked list). $\Delta \phi = 1$ since one node is added to $B$ as a new tree. Amortized cost: $O(1) + O(1) = O(1)$.\\\\
\textbf{find-min}: $O(1)$ actual (since $A$ is a lazy binomial heap, we can use $O(1)$ here as ``actual'' even though part of it is amortized). $\Delta \phi = 0$ since no trees are added and $A$ is not changed.\\\\
\textbf{meld}$(B,pq_2)$: $O(1)$ actual (add to $A$, subtract counter). $\Delta \phi = 1$ since one node is added to $A$. \\\\
\textbf{add-to-all}$(s)$: $O(1)$ actual (updating $c$). $\Delta \phi = 0$ since no trees are added.\\\\
\textbf{extract-min}: \\\\ Actual $O(1)$ (find minimum) + $O(\log n)$ (exposing children on extract and adding d to children aux values) + $O(\text{\# trees in} B)$ (merge trees) + $O(\text{\# nodes in} A)$ (merge melded heaps). $\Delta \phi = -T_A-T_B + O(\log n)$ (remove $T_A$ nodes from $A$, remove $T_B$ trees from $B$, create O($\log n$) of them after merging). \\\\ Amortized: $O(1) + O(\log n) + O(T_A) + O(T_B) + O(1) \cdot (-T_A - T_B + O(\log n)) = O(1) + O(\log n) + O(\log n) = O(\log n)$
\end{proof}

\end{document}



%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
